Profiles of Wind Direction and Speed in Turbulent Ekman Flow

### 2021-09-29 write1/manuscript.tex
comments on current draft of part 2; worked on profile description in Ekman layer

@@

requires sufficient scale separation in the simulations to not only quantify the effect of
the Reynolds number on low-order flow statistics, but also assess the corresponding rate-of-change

@@

While the simulations available give confidence in a first-order representation of the turbulent

@@

Velocity profiles in inner and units (Fig.~\ref{fig:profiles}a,b) are in accordance

@@

the corresponding profiles $U^{\alpha_\star}(z^-)/u_\star$ agree closely,
irrespective of $\RE$ which illustrates the
inner--outer scale-duality in this external flow. 

@@

It is noteworthy that also in the outer layer of the flow, $u_\star$ (not the magnitude the

@@

When normalized in outer units, the profiles of spanwise velocity $V^{\alpha_\star}(z^-)/u_\star$

@@

@@

appears to converge to an $\RE$-independent limit within the range of scale separation considered
here.

@@

  \nu_E^-=\nu_E^+  \frac{Z_{\star}}{\delta^+} = \nu_E \frac{1}{\nu} \frac{G}{u_\star}\frac{\nu}{u_\star \delta}  = \nu _E \frac{Gf}{u_\star^3}. 

@@

Substantial variation of the profiles is, however observed below and above this minimum for different $\RE$ which illustrates that
this normalization is probably not generally appropriate across the outer layer.

@@

In the upper part of the outer layer, around $y^+\approx 1000$, no clearly visible signature of the

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
\label{sec:profiles} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We now turn to the formulation of a general velocity profile that is fully determined by the only
parameter of the idealized problem, namely a Reynolds number representing the scale separation or
geometric size of the problem.
%
This precludes a drag law with wherewith we begin this section (\ref{sec:drag-law}).
Based on the scaling arguments put forward in Sec.~\ref{sec:scaling}, we then develop
a formulation of the wind vector in the Ekman layer (Sec~\ref{sec:ekman}) and separately
for the streamwise (Sec.~\ref{sec:streamwise}) and spanwise components (Sec.~\ref{sec:spanwise})
in the overlap and inner regions. 
%

\end{itemize}
\begin{figure}
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of geostrophic drag, $Z_\star$, and surface veering, $\alpha_\star$, with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data
    \label{fig:drag_law}}
\end{figure}
% 
\subsection{Drag-Law}
\label{sec:drag-law}
A drag-law for Ekman flow determines two key parameters of the flow as a function of Reynolds number alone.
%
These are, first, the modulus of the surface friction normalized by the geostrophic wind forcing, $u_\star$ and,
second, the direction of surface shear stress, $\alpha_\star$, also termed wind veering. 
%
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system for analysis
(namely that aligned with the surface shear stress) is a priori unknown. 
%
We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
%
\par
%
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
\begin{equation}
  u_\star= 4\log(\mathbf{Re}_D))-8
\end{equation}
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one, at least
for problems with a scale separation on the order that is relevant to geophysical problems ($Re_D<10^{8}$).  
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection{Profile in the Ekman layer}
\label{sec:ekman}
%
Formulations for the outer layer that take into account the rotation
(and thus deviation from the channel-flow analogy)
need to be matched to the framework of surface similarity.
%
A smooth transition from the inner layer to the Ekman layer, where the wind is characterized by a turning
of its mean direction, is not easily achieved. 
%
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when coming approaching the Ekman layer from below.
%
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to,
we understand this as the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.
%
The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile.
%
The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift
in reference height.
%
\begin{figure}
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
  \caption{} 
\end{figure}

\subsection{Streamwise Velocity Component} 
For the streamwise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

@@

>>>>

  \label{eqn:alpha_def} 

@@

requires sufficient scale separation in the simulations, not only to quantify the effect of
the Reynolds number on low-order flow statistics, but also to assess the corresponding rate-of-change

@@

While the simulations previously available (gray shading in Tab.~\ref{tab:sim-setup})
give confidence in a first-order representation of the turbulent

@@

Velocity profiles in inner and outer units (Fig.~\ref{fig:profiles}a,b) are in accordance

@@

the corresponding profiles of the velocity deficit  $(U^{\alpha_\star}(z^-) - G_{1}^{\alpha} ) /u_\star$
agree closely, irrespective of $\RE$.

@@

This illustrates the inner--outer scale-duality in this external flow with inner scaling being
appropriate in the inner layer and outer scaling in the outer layer. 
Also in the outer layer of the flow, $u_\star$ (not the magnitude the

@@

When normalized in outer units, the deficit profiles of spanwise velocity
$(V^{\alpha_\star}(z^-)-G_{2}^{\alpha})/u_\star$

@@

The value of $V^{\alpha_\star+}(z^-)-G_{2}^{\alpha}$ is sensitive to the wind veering for
$z\rightarrow 0$ as---for use of the shear-aligned component---it has to approach the value
$-G_2^\alpha=|G|\sin\alpha\ne0$ in view of the no-slip boundary condition.  

@@

converges to an $\RE$-independent limit within the range of scale separation considered
here, i.e.
  G_{2}^{\alpha}/u_\star = Z_\star \sin\alpha \rightarrow \text{const. for\ }\RE\rightarrow \infty, 
which has indeed already been used by \cite{spalart:JFM1989}. 

@@

  \nu_E^-=\nu_E^+  \frac{1}{\delta^+} = \nu_E \frac{1}{\nu} \frac{\nu}{u_\star \delta}  = \nu _E \frac{f}{u_\star^2}.
  \label{eqn:evisc_outer_scaling}

@@

Substantial variation of the profiles is, however observed below and above this minimum for
different $\RE$ which illustrates that this normalization is probably not generally appropriate
across the outer layer.

@@

In the upper part of the outer layer, around $y^+\approx 1000$, no clear signature of the

@@

\section{Towards a Universal Velocity Profile for the Turbulent Ekman Layer}
\label{sec:profiles} 
We now turn to the formulation of a general velocity profile that is fully determined by the only
parameter of the idealized problem, namely a Reynolds number representing the scale separation or
geometric size of the problem.
This precludes a drag law wherewith we begin this section (\ref{sec:drag-law}).
Based on the scaling arguments put forward in Sec.~\ref{sec:scaling}, we then develop
a formulation of the wind vector in the Ekman layer (Sec~\ref{sec:ekman}) and separately
for the streamwise (Sec.~\ref{sec:streamwise}) and spanwise (Sec.~\ref{sec:spanwise}) components
in the overlap and inner regions. 
  \centerline{\includegraphics[width=0.7\textwidth]{../plot/ustar_alpha.pdf}}
  \caption{Variation of geostrophic drag, $Z_\star$, and surface veering, $\alpha_\star$, with Reynolds number according to
    the theory by Spalart et al. (1989) and as estimated from DNS data
    \label{fig:drag_law}}
\subsection{Drag-Law}
\label{sec:drag-law}
A drag-law for Ekman flow determines---as a function of Reynolds number alone---the surface drag.
This can be formulated in terms of the modulus
of the surface friction normalized by the geostrophic wind forcing, $u_\star$ (Eq.~(\ref{eqn:ustar_def}))
and, the direction of surface shear stress, $\alpha_\star$ (Eq.~(\ref{eqn:alpha_def})),
also termed wind veering. 
A non-zero veering of the wind is a rather special case in comparison with most turbulent flows conisdered in an
engineering context, and it confronts us with a situation where the most appropriate coordinate system
for analysis (namely that aligned with the surface shear stress) is a priori unknown. 
We compare our DNS data against a semi-empirical drag-law based on integral consideration \cite{spalart:JFM1989}
and find, as demonstrated in previous work \cite{ansorge:BM2014}, excellent agreement in the range
 $400<\mathbf{Re}<1600$, representing a factor of 16 in variation of viscosity. 
We also find that the solution of the transient equation involved in estimation of $u_\star$ for a given Reynolds number
$Re_D$ is approximated reasonably by the formulation
  u_\star= 4\log(\mathbf{Re}_D))-8
which quantifies the 'weak' dependence of $u_\star$ on the Reynolds number as an approximately logarithmic one, at least
for problems with a scale separation on the order that is relevant to geophysical problems ($Re_D<10^{8}$).  
\subsection{Profile in the Ekman layer}
\label{sec:ekman}
Formulations for the outer layer that take into account the rotation
(and thus deviation from the channel-flow analogy)
need to be matched to the framework of surface similarity.
A smooth transition from the inner layer to the Ekman layer, where the wind is characterized by a turning
of its mean direction, is not easily achieved. 
\cite{optis:BM2014}, for instance, define an ``effective geostrophic wind vector that has the same magnitude of the
observed surface geostrophic wind and is rotated by the angle $\alpha$ [their nomenclature]'' to overcome the
unsteady transition when approaching the Ekman layer from below.
While such rotation of the wind vector is \emph{a posteriori} justified by the observational data that the model outcomes are compared to, understand this as the manifestation of a mismatch between the theoretical treatment of the inner and outer layer of
turbulent Ekman flow.
The classic text-book solution of Ekman flow makes use of the physical boundary conditions (BCs) for the atmospheric
boundary layer (no-slip at the bottom and geostrophic wind in the free atmosphere).
These two boundary conditions eliminate one mode of the analytical solution and determine the magnitude of
the spiral.
In doing so, one has to assume that the solution is appropriate across the entire ABL, which is not the case:
the dynamics put forth by \cite{ekman:AMA1905} are not appropriate in the surface layer of the ABL, where
the logarithmic, buffer, and viscous sublayers are found. 
We to use a virtual boundary condition instead that takes into account the increased gradients in vicinity of
the surface and matches the magnitude of the Ekman spiral to the values observed in our DNS:
  G^{-1}\left(\begin{array}{c} 
    U_\mathrm{ek}\\ 
    V_\mathrm{ek}
  \end{array}\right)  &=& \left(\begin{array}{c} 1 \\ 0 \end{array}\right) 
  + a_\mathrm{ek} e^{-z_\mathrm{ek}} \left(\begin{array}{r}
    -\cos z_\mathrm{ek} \\ 
    \sin z_\mathrm{ek}
  \end{array}\right)  \\
  z_\mathrm{ek} &=& \delta_{ek} (y^- -s_{ek}). 
This solution is characterized by three parameters,
(i) an Ekman-layer depth scale $\delta_\mathrm{ek}$,
(ii) the magnitude parameter of the spiral $a_{ek}$, and
(iii) a zero-crossing point for the velocity $s_{ek}$.
\begin{enumerate} 
\item[(i)] The Ekman-layer depth scale $\delta_\mathrm{ek}$ is fundamentally defined by the eddy diffusivity.
  However, we have seen in Section \ref{sec:scaling} that a characteristic value for the eddy diffusivity
  is not easily obtained for its strong dependence on the Reynolds number and distance from the surface.
  We therefore resort to the physical manifestation of the eddy diffuysivity in an Ekman layer, and use the
  boundary layer depth $\delta_\mathrm{ek} = 0.66 \delta \times 2\pi$.
  For the relation $\delta_\mathrm{ek}=\sqrt{2\nu_\mathrm{ek}/f}$, this yields
  $\nu_\mathrm{ek} \propto u_\star^2/f$ in accordance with the observations in
  Sec.~\ref{sec:scaling} (Eq.~\ref{eqn:evisc_outer_scaling}). 
\item[(ii)] The magnitude parameter of the spiral primarily defines the super-geostrophic maximum of the
  wind profile aloft the logarithmic layer. Our simulations suggest this maximum of the velocity deficit
  remains constant when normalized by $u_\star$ as shown in Fig.~\ref{fig:profiles_outer}.
  The numerical value of $a_\mathrm{ek}$ is estimated from visual comparison, and we find $a_{ek}=8.4u_\star$;
  while this appears rather large, it is pre-multiplied by $e^{-z_\mathrm{ek}}$ which has already decreased
  to $\mathcal{O}(0.1)$ at the height of this maximum.
  This choice ascertains that the velocity deficits $U/u_\star-Z_\star$ and $V/u_\star-Z_\star$ do not
  depend on the velocity scale $u_\star$, but only on $G$ as
    U_\mathrm{ek}/u_\star-Z_\star \propto a_\mathrm{ek} Z_\star = 8.4G.  \notag
\item[(iii)] While $\delta_\mathrm{ek}$ sets the thickness across which the wind veering takes place,
  the zero-crossing height $s_\mathrm{ek}$ defines an offset.
  Physically, this offset can be understood as the height at which the surface was located assuming
  a perfect Ekman flow down to the surface.
  As this is not the case, and gradients are steeper in the highly turbulent boundary layer flow encountered
  when approaching the surface, the offset is smaller than zero (the fully turbulent boundary layer is
  actually thinner than an Ekman layer would be).
  From our DNS data, we estimate $s_\mathrm{ek}=-0.12\delta$. 
\end{enumerate} 
\todo{Paragraph and figure on idealized Ekman profile}
The outer layer of Ekman flow is characterized by a turning of the wind velocity and the super-geostrophic maximum
that is sustained by momentum convergence at the inflection point of the velocity profile.
The super-geostrophic maximum of streamwise velocity and a secondary minimum aloft the bulk-turbulent part of
the boundary layer are well-described by a classic Ekman spiral with adapted boundary conditions and a shift
in reference height.
  \includegraphics[width=0.5\textwidth]{../plot/outer_layer_u.pdf}\includegraphics[width=0.5\textwidth]{../plot/outer_layer_w.pdf}
    \label{fig:profiles_outer} 

\subsection{Streamwise Velocity Component}
\label{sec:streamwise} 
For the streamwise velocity profile (that in non-rotating flows due to the geometry is always aligned with the surface shear stress),
well-established theories exist for various regimes according to their distance from the wall and the relative role of
viscosity, turbulence and interaction with the outer region of the flow with the logarithmic law for the mean velocity as a central anchor point.
%
\par
%
\begin{subequations}

@@

\label{sec:spanwise} 
